SecFFT: Safeguarding Federated Fine-tuning for Large Vision Language Models against Covert Backdoor Attacks in IoRT Networks

### 2024-11-29 main-v3(final).tex
before.final: modified by zz
Signed-off-by: LetMeFly666 <814114971@qq.com>

%如图\ref{fig:scenario}所示，考虑在机器人物联网（IoRT）网络中对视觉大模型进行联邦微调的场景下，整个系统模型包括多个机器人节点和一个中心服务器。每轮，每个机器人节点在本地对视觉大模型进行微调并只上传微调部分的参数更新，随后中心服务器对所有更新进行聚合，然后下发聚合后的参数，开启新一轮的联邦微调。以下是系统模型的详细描述。
As illustrated in Figure \ref{fig1:scene}, \textcolor{red}{consider the scenario of federated fine-tuning (FFT) for large vision language models (LVLMs) in an Internet of Robotic Things (IoRT) network. The entire system model consists of multiple robot nodes and a central server. In each round, each robot node locally fine-tunes the vision model and only uploads the updates of the fine-tuned part. The central server then aggregates all updates and distributes the aggregated parameters to initiate a new round of federated fine-tuning. Below is a detailed description of the system model.}

%\textbf{参与者和数据集}：假设系统模型中存在 $N$ 个机器人节点，每个节点记作 $R_i$, 其中$i \in \{1, 2, \ldots, N\}$。 每个机器人节点 $R_i$ 拥有本地数据集 $D_i = \{x_{i,j}, y_{i,j}\}_{j=1}^{|D_i|}$，其中 $x_{i,j}$ 表示训练样本，$y_{i,j}$ 表示对应的标签数据，$|D_i|$ 为数据集的大小，所有节点本地数据集的并集$D = \bigcup_{i=1}^{N} D_i$构成整体数据集。
\textbf{Participants and Datasets}:
\textcolor{red}{Assume there are $N$ robot nodes in the system model, with each node denoted as $R_i$, where $i \in {1, 2, \ldots, N}$. Each robot node $R_i$ possesses a local dataset $D_i = {x_{i,j}, y_{i,j}}{j=1}^{|D_i|}$, where $x{i,j}$ represents a training sample, $y_{i,j}$ represents the corresponding label, and $|D_i|$ denotes the size of the dataset. The union of all local datasets across the nodes, $D = \bigcup_{i=1}^{N} D_i$, constitutes the overall dataset.}

@@

%\textbf{LoRA}： 在LoRA(Low-Rank Adaptation) \cite{lora}方法中，每个参数矩阵 $W$ 被表示为：
\textbf{Fine-tuning}:
Taking Low-Rank Adaptation (LoRA) \cite{lora} as an example, model parameter matrix $W$ can be expressed as:
\begin{equation}
    W = W_0 + \Delta W, \quad \Delta W = BA,
\end{equation}
    %其中，$W_0$ 为模型的原始权重矩阵，$\Delta W$ 是低秩增量矩阵，$B$ 和 $A$ 是低秩分解矩阵，$r$ 为矩阵的秩。在对视觉大模型进行联邦微调的场景中，全局模型原始参数为$\Theta_p = \sum {W_0}$，而LoRA引入的增量参数为$\theta_p = \sum {W}$，因此全局模型的整个参数集为$\Theta = \Theta_p + \theta_g$。在整个联邦微调过程中，仅对全局模型的增量参数$\theta_g$ 进行微调训练，保持原始参数$\Theta_p$不变。
\textcolor{red}{where $W_0$ represents the original weight matrix of the model, $\Delta W$ denotes the low-rank increment matrix, $B$ and $A$ are the low-rank decomposition matrices, and $r$ is the rank of the matrix. In the scenario of FFT for LVLMs, the fixed parameters of the global model are given by $\Theta_p = \sum {W_0}$, and the incremental parameters introduced by LoRA are denoted as $\theta_g = \sum {W}$. Therefore, the entire parameter of the global model is $\Theta = \Theta_p + \theta_g$. Throughout the FFT process, only the parameters $\theta_g$ of the global model are fine-tuned, while the parameters $\Theta_p$ remain unchanged.}

@@

%\item \textbf{联邦聚合}：当中心服务器接收到所有机器人节点的本地更新$\triangledown \theta_i^t$后，利用相应的聚合算法进行聚合得到新一轮的全局增量参数$\theta_g^t$：
\textbf{Federated Aggregation}:
\textcolor{red}{After the central server receives the updates $\triangledown \theta_i^t$ from all robot nodes, it applies the appropriate aggregation algorithm to generate the next round's global incremental parameters $\theta_g^t$:
% The central server collects the local updates $\triangledown \theta_i^t$ from each robot node $R_i$ and aggregates them using a specified aggregation algorithm to update the global incremental parameters $\theta_g^t$:   
\begin{equation}
    \theta_g^t = \theta_g^{t-1} + Agg(\triangledown \theta_1^t, \triangledown \theta_2^t, \dots, \triangledown \theta_n^t),
\end{equation}
%其中，$Agg$表示中心服务器使用到的聚合算法，例如FedAvg、FLTrust\cite{cao2020fltrust}、RoseAgg\cite{yang2024roseagg}等，$\theta_g^t$ 是第 $t$ 轮更新的全局增量参数，更新后的全局增量参数 $\theta_g^t$ 将在下一轮训练前下发给所有机器人节点，用于进一步的本地微调。
where $Agg$ denotes the aggregation algorithm employed by the central server, such as FedAvg, FLTrust \cite{cao2020fltrust}, or RoseAgg \cite{yang2024roseagg}.}

\section{Threat Model}
% 在对视觉语言模型进行联邦微调的场景中，我们假设存在一个或者多个由攻击者操纵的恶意节点，并且恶意节点的数量小于全部参与节点的一半。攻击者的目标是对全局模型$\Theta$ 进行后门植入，使得模型能够将任何带有预定义触发器 $\delta$ 的测试输入分类为目标类别，即$\Theta(x+\delta) = y'$，其中$x$代表正常输入，$y'$表示目标类别。同时尽可能保持对干净输入的分类准确性，即$\Theta(x) = y$，其中$y$为输出$x$的原始类别。
\textcolor{red}{In the scenario of FFT for LVLMs, we assume the presence of one or more malicious nodes controlled by attackers, with the number of malicious nodes being less than half of all participating nodes.}

@@

%Attacker's Ability: 攻击者对恶意节点的本地训练过程具有完全的控制权，意味着攻击者可以任意操控本地数据、触发器模式、优化策略和本地更新。此外，攻击者可以采用“智能”攻击策略，这种策略的特点是攻击效果不会在单轮训练中显现，而是通过多轮训练的累积效应逐步将后门植入全局模型中。这种场景在实际中是可能存在的，因为中央服务器只能接收到客户端上传的训练后模型更新，而无法知晓每个客户端的具体训练过程。然而，攻击者无法对中央服务器的操作进行干扰，例如，无法更改全局模型的聚合规则或篡改其他良性客户端的本地更新。
\textbf{Attacker's Ability:} 
\textcolor{red}{The attacker has complete control over the local training process of the malicious robot nodes, meaning they can manipulate local data, trigger patterns, optimization strategies, and local updates at will. Additionally, the attacker may adopt a stealthy attack strategy, where the backdoor is not implanted in a single training round but gradually embedded into the global model through the cumulative effect of multiple rounds of training. This scenario is plausible in practice, as the central server only receives the updates uploaded by robot nodes and has no insight into the specific training process of each node. However, the attacker cannot interfere with the aggregation process of the central server (i.e., altering the global model aggregation rules) or tamper with the local updates of benign nodes.}

@@

\subsection{Instantaneous Attack Behavior Perception}
\begin{figure}[!t]
    \centering
    \includegraphics[width=1\linewidth]{figures/secfft-1.pdf}
    \caption{SecFFT's instantaneous attack behavior perception module with frequency-domain deep outlier feature semantics extraction}
    \label{fig2:instant}
\end{figure}
%^\textbf{Motivation:} 目前的防御策略主要集中于利用攻击的“浅层语义特征”来进行恶意节点识别，所谓浅层语义是指模型上传更新的方向\cite{foolsgold}，范数\ref{sun2019can}以及几何中心\cite{pillutla2022robust}等几何或者统计特性。然而，这些基于浅层特征的防御方法在高维空间中存在显著的局限性。首先，这些防御方式只能在特定假设条件下对某些类型的攻击起作用。例如，余弦距离能够检测出具有较大方向偏差的恶意更新，而欧几里得距离则用于检测通过扩大$L_2$范数影响全局模型的恶意更新。同时最新的研究表明\cite{huang2023multi}，基于欧几里得距离等传统的距离度量方法在高维空间中无法有效区分恶意和良性更新，因为这些方法往往受到维度灾难的影响。此外攻击策略也在不停的演变，现在的攻击者已经能够通过精心调整，使恶意更新在方向、范数等浅层特征上与良性更新相似，从而巧妙地避开这些基于浅层语义特征的防御措施，最终在全局模型中植入后门。但尽管攻击者的攻击策略变得越来越复杂，值得注意的是，后门攻击的本质在于让模型建立起预定义触发器和目标之间的联系，因此无论攻击者如何调整自己的攻击方式或策略，其上传的模型更新相较于良性用户始终在某种分布上具有偏移。
\textbf{Motivation:} 
Current defense strategies primarily focus on identifying malicious nodes by leveraging shallow semantic features of the attack, such as the direction \cite{foolsgold}, magnitude \cite{sun2019can}, and geometric or statistical properties like geometric centers \cite{pillutla2022robust} of updates. However, these defense methods based on shallow semantic features face significant limitations in high-dimensional spaces. First, these methods are only effective against certain types of attacks under specific assumptions. For example, cosine distance can detect malicious updates with large directional deviations, while Euclidean distance is used to detect malicious updates that affect the global model by increasing the $L_2$ norm. Furthermore, recent studies \cite{huang2023multi} have shown that traditional distance metrics, such as Euclidean distance, cannot effectively distinguish between malicious and benign updates in high-dimensional spaces due to the curse of dimensionality. Additionally, attack strategies are constantly evolving, and attackers can now carefully adjust their malicious updates to resemble benign updates in terms of direction, magnitude, and other shallow features. This allows them to bypass defenses based on shallow semantic features and eventually implant backdoors into the global model. \textcolor{red}{Despite the increasingly sophisticated strategies employed by attackers, it is important to note that the essence of backdoor attacks lies in establishing a link between the predefined trigger and the target within the model. Thus, regardless of how attackers adjust their attack strategies, their updates will always exhibit some distributional shift compared to benign nodes.}

@@

%\textbf{Overview:} 在神经网络中，每个权重代表两个神经元之间连接的强度，权重分布及其相关的能量在训练过程中会经历动态演变，以缩小模型预测结果与真实目标类别之间的差距。当后门攻击迫使神经网络模型学习一种在正常数据中不存在的关联模式时，根据后门触发器的设计和实现方式，这种模式可能是微妙的，也可能是相当明显的，但无论如何，它都代表了与模型从正常数据中学习到的模式的偏离。因此，为解决基于浅层语义特征的防御方法在高维空间中面临的局限性，我们从后门攻击的本质出发\cite{zeng2021rethinking}，提出了一种基于深层语义特征的防御策略，重点分析模型更新在频域空间上的差异。通过将本地模型更新转换到频域，我们能够更准确地捕捉恶意更新在低频成分上的异常表现，这些异常通常难以被传统的浅层特征防御方法发现。通过对频域上的相应特征进行分析，我们可以有效地区分出恶意节点与良性节点，从而增强联邦学习系统在面对复杂后门攻击时的鲁棒性和检测精度。
\textbf{Overview:} 
In neural networks, each weight represents the strength of the connection between two neurons, and the distribution of these weights, along with their associated energy, undergoes dynamic evolution during training to reduce the gap between the model's predicted results and the true classes. nonetheless, When a backdoor attack forces the neural network to learn an association that does not exist in normal data, which may be subtle or quite obvious depending on the design and implementation of the backdoor trigger, it invariably represents a deviation from the association learned by the model from normal data. \textcolor{red}{To address the limitations faced by defense methods based on shallow semantic features in high-dimensional spaces, we propose a new defense strategy rooted in the essence of backdoor attacks \cite{zeng2021rethinking}, which leverages deep semantic features by focusing on the differences in model updates in the frequency domain. According to our experimental results, by transforming local model updates into the frequency domain, we can more accurately capture the abnormal behavior of malicious updates which is typically difficult to detect using traditional shallow feature-based defenses in low-frequency components. By analyzing the relevant features in the frequency domain, we can effectively distinguish between malicious and benign nodes, thereby enhancing the robustness and accuracy of federated learning in the face of sophisticated backdoor attacks.}

@@

\subsubsection{Frequency domain transformation}
%受高维空间中维度灾难的影响，传统的基于距离度量的方法已无法有效识别恶意节点。相反，将模型更新转换到频率域后，其差异将主要集中在低频成分上。因此，专注于低频成分的分析，在提升识别准确率的同时，将显著的降低计算开销。假设每个节点上传的更新为 \(\triangledown \theta_i^t \in \mathbb{R}^d\)，其中 \(i\) 表示机器人节点，\(t\) 表示通信轮次，\(d\) 为本地模型上传更新的维度。对于每个更新 \(\triangledown \theta_i^t\)，对其进行归一化，然后通过一维的DCT-II\cite{ahmed1974discrete}变化将其转换为对应的频率分布，并提取$m$个低频分量，例如5000，得到低频向量\(G_i^t\)。
Due to the curse of dimensionality in high-dimensional spaces, traditional methods based on shallow semantic
feature can no longer effectively identify malicious nodes. In contrast, once model updates are transformed into the frequency domain, their differences are primarily concentrated in the low-frequency components. Focusing on the analysis of these low-frequency components not only improves identification accuracy but also significantly reduces computational overhead. \textcolor{red}{For each update \(\triangledown \theta_i^t\), we apply one-dimensional DCT-II \cite{ahmed1974discrete} to obtain its corresponding frequency distribution. Then we extract $m$ low-frequency components\cite{fereidooni2023freqfed}, such as 5000, yielding the low-frequency vector $G_i^t$.}

@@

\subsubsection{Clean ingredient extraction}
% 在提取到节点更新对应的低频向量\(G_i^t\)后，我们使用流行的聚类算法例如HDBSCAN\cite{mcinnes2017hdbscan}对这些更新进行聚类，得到多个簇$C_0^t, C_1^t, \ldots, C_{\kappa}^t$，其中$\kappa$表示簇的数量。在大多数联邦学习场景中，良性节点的数量通常多于恶意节点的数量，且良性节点的更新通常更接近全局模型的方向并表现出更高的聚集性。基于这一观察，我们选择节点数量最多的簇$C_{max} = \{R_{i_0}, R_{i_1}, \ldots, R_{i_n}\}$作为主要分析对象。为了进一步提取该簇的主要特征或模式，我们对其中所有节点组成的矩阵进行奇异值分解\cite{wang2023scfl}。由于，SVD分解的左奇异向量$\tilde{G}^t$对应于数据的主要方向，因此，$\tilde{G}^t$可以作为良性节点在低频域内的主要特征。
After extracting the corresponding low-frequency vectors \(G_i^t\) from updates, we employ popular clustering algorithms such as HDBSCAN \cite{mcinnes2017hdbscan} to cluster these vectors, resulting in multiple clusters \(C_0^t, C_1^t, \ldots, C_{\kappa}^t\), where \(\kappa\) denotes the number of clusters. In most federated learning scenarios, the number of benign nodes typically exceeds that of malicious nodes, which is consistent with the assumptions outlined in our threat model. Moreover, the updates from benign nodes are generally closer to the direction of the global model and exhibit a higher degree of aggregation. This behavior is also evident in the frequency domain, where it becomes even more pronounced. 
\textcolor{red}{Based on this observation, we select the largest cluster \(C^t_{max} = \{G^t_{i_0}, G^t_{i_1}, \ldots, G^t_{i_n}\}\) as the primary subject for analysis. To further extract the main characteristics of this cluster, we perform singular value decomposition (SVD)\cite{wang2023scfl} on the matrix $H^t = (G^t_{i_0}, G^t_{i_1}, \ldots, G^t_{i_n}) \in \mathbb{R}^{m \times n}$ formed by all nodes in this cluster. Since the left singular vectors of the SVD, denoted as \(\tilde{G}^t\), represent the principal directions of these vectors, \(\tilde{G}^t\) can be regarded as the principal ingredient of benign nodes in the low-frequency domain.}

@@

\textcolor{red}{
\begin{equation}
Chi^t_i = \sqrt{ \sum_{k=0}^{m-1}\frac{(G_i^t[k] - {\tilde G}^t[k])^2}{{|\tilde G}^t[k]| + \epsilon} }
\label{equation:KL}
\end{equation}}
% 其中 \(m\) 是低频向量的维度，\(\epsilon\) 是一个小的常数，用于避免分母为零。相比于传统的欧氏距离，卡方距离根据两个向量对应分量的大小进行加权，使得特征值较大的部分对距离的影响更大，从而能够更准确地反映低频向量中重要特征之间的差异，同时忽略不重要的噪声成分。
where \(G_i^t[k]\) denotes the k-th element of vector \(G_i\) and \(\epsilon\) is a small constant to avoid division by zero.

\textcolor{red}{
Chi-square distance is usually used to measure whether a certain observed distribution conforms to a typical theoretical distribution, which is very consistent with our method because we use the calculated clean ingredient as a criterion for measuring benign and malicious nodes. Compared with the traditional Euclidean distance, the chi-square distance assigns weights according to the clean ingredient, giving greater importance to smaller components. On the other hand, cosine similarity measures the angle between two vectors and is usually used to evaluate their directional consistency, ignoring magnitude differences. However, in our method, we are concerned with the degree of deviation of each user's low-frequency vector $G_i^t$ from the clean ingredient $\tilde{G}^t$, including magnitude differences, so the chi-square distance is more suitable for capturing hidden malicious attacks than cosine similarity.
}
% 为什么要用卡方距离，不用欧式距离？实验论证？

@@

\textcolor{red}{In our proposed framework, the concept of the "intention point" serves as a pivotal element in recognizing the underlying motives of malicious nodes during the federated fine-tuning process. The intention point is defined as the central position within a high-dimensional space, representing the aggregated gradient updates of a node across multiple training rounds. This point encapsulates the node's intended direction for influencing the global model. To construct the intention point, we utilize a Minimum Enclosing Hypersphere approach, which encompasses a proportion of the node’s gradient updates. The radius of this hypersphere reflects the reliability of the intention point, effectively quantifying the node's consistency in updating the model. By analyzing these intention points, our method can discern the behaviors of malicious nodes from benign ones, enhancing the overall security and robustness of the federated learning system.}

>>>>

%如图\ref{fig:scenario}所示，考虑在机器人物联网（IoRT）网络中对视觉大模型进行联邦微调的场景下，整个系统模型包括多个机器人节点和一个中心服务器。每轮，每个机器人节点在本地对视觉大模型进行微调并只上传微调部分的参数更新，随后中心服务器对所有更新进行聚合，然后下发聚合后的参数，开启新一轮的联邦微调。以下是系统模型的详细描述。
As illustrated in Figure \ref{fig1:scene}, consider the scenario of federated fine-tuning (FFT) for large vision language models (LVLMs) in an Internet of Robotic Things (IoRT) network. The entire system model consists of multiple robot nodes and a central server. In each round, each robot node locally fine-tunes the vision model and only uploads the updates of the fine-tuned part. The central server then aggregates all updates and distributes the aggregated parameters to initiate a new round of federated fine-tuning. Below is a detailed description of the system model.

%\textbf{参与者和数据集}：假设系统模型中存在 $N$ 个机器人节点，每个节点记作 $R_i$, 其中$i \in \{1, 2, \ldots, N\}$。 每个机器人节点 $R_i$ 拥有本地数据集 $D_i = \{x_{i,j}, y_{i,j}\}_{j=1}^{|D_i|}$，其中 $x_{i,j}$ 表示训练样本，$y_{i,j}$ 表示对应的标签数据，$|D_i|$ 为数据集的大小，所有节点本地数据集的并集$D = \bigcup_{i=1}^{N} D_i$构成整体数据集。
\textbf{Participants and Datasets}: Assume there are $N$ robot nodes in the system model, with each node denoted as $R_i$, where $i \in {1, 2, \ldots, N}$. Each robot node $R_i$ possesses a local dataset $D_i = {x_{i,j}, y_{i,j}}{j=1}^{|D_i|}$, where $x{i,j}$ represents a training sample, $y_{i,j}$ represents the corresponding label, and $|D_i|$ denotes the size of the dataset. The union of all local datasets across the nodes, $D = \bigcup_{i=1}^{N} D_i$, constitutes the overall dataset.

@@

%\textbf{LoRA}： 在LoRA(Low-Rank Adaptation) \cite{lora}方法中，每个参数矩阵 $W$ 被表示为：
\textbf{Fine-tuning}:
Taking Low-Rank Adaptation (LoRA) \cite{lora} as an example, model parameter matrix $W$ can be expressed as:
\begin{equation}
    W = W_0 + \Delta W, \quad \Delta W = BA,
\end{equation}
    %其中，$W_0$ 为模型的原始权重矩阵，$\Delta W$ 是低秩增量矩阵，$B$ 和 $A$ 是低秩分解矩阵，$r$ 为矩阵的秩。在对视觉大模型进行联邦微调的场景中，全局模型原始参数为$\Theta_p = \sum {W_0}$，而LoRA引入的增量参数为$\theta_p = \sum {W}$，因此全局模型的整个参数集为$\Theta = \Theta_p + \theta_g$。在整个联邦微调过程中，仅对全局模型的增量参数$\theta_g$ 进行微调训练，保持原始参数$\Theta_p$不变。
where $W_0$ represents the original weight matrix of the model, $\Delta W$ denotes the low-rank increment matrix, $B$ and $A$ are the low-rank decomposition matrices, and $r$ is the rank of the matrix. In the scenario of FFT for LVLMs, the fixed parameters of the global model are given by $\Theta_p = \sum {W_0}$, and the incremental parameters introduced by LoRA are denoted as $\theta_g = \sum {W}$. Therefore, the entire parameter of the global model is $\Theta = \Theta_p + \theta_g$. Throughout the FFT process, only the parameters $\theta_g$ of the global model are fine-tuned, while the parameters $\Theta_p$ remain unchanged.

@@

%\item \textbf{联邦聚合}：当中心服务器接收到所有机器人节点的本地更新$\triangledown \theta_i^t$后，利用相应的聚合算法进行聚合得到新一轮的全局增量参数$\theta_g^t$：
\textbf{Federated Aggregation}: After the central server receives the updates $\triangledown \theta_i^t$ from all robot nodes, it applies the appropriate aggregation algorithm to generate the next round's global incremental parameters $\theta_g^t$:
% The central server collects the local updates $\triangledown \theta_i^t$ from each robot node $R_i$ and aggregates them using a specified aggregation algorithm to update the global incremental parameters $\theta_g^t$:   
\begin{equation}
    \theta_g^t = \theta_g^{t-1} + Agg(\triangledown \theta_1^t, \triangledown \theta_2^t, \dots, \triangledown \theta_n^t),
\end{equation}
%其中，$Agg$表示中心服务器使用到的聚合算法，例如FedAvg、FLTrust\cite{cao2020fltrust}、RoseAgg\cite{yang2024roseagg}等，$\theta_g^t$ 是第 $t$ 轮更新的全局增量参数，更新后的全局增量参数 $\theta_g^t$ 将在下一轮训练前下发给所有机器人节点，用于进一步的本地微调。
where $Agg$ denotes the aggregation algorithm employed by the central server, such as FedAvg, FLTrust \cite{cao2020fltrust}, or RoseAgg \cite{yang2024roseagg}.

\section{Threat Model}
% 在对视觉语言模型进行联邦微调的场景中，我们假设存在一个或者多个由攻击者操纵的恶意节点，并且恶意节点的数量小于全部参与节点的一半。攻击者的目标是对全局模型$\Theta$ 进行后门植入，使得模型能够将任何带有预定义触发器 $\delta$ 的测试输入分类为目标类别，即$\Theta(x+\delta) = y'$，其中$x$代表正常输入，$y'$表示目标类别。同时尽可能保持对干净输入的分类准确性，即$\Theta(x) = y$，其中$y$为输出$x$的原始类别。
In the scenario of FFT for LVLMs, we assume the presence of one or more malicious nodes controlled by attackers, with the number of malicious nodes being less than half of all participating nodes.

@@

%Attacker's Ability: 攻击者对恶意节点的本地训练过程具有完全的控制权，意味着攻击者可以任意操控本地数据、触发器模式、优化策略和本地更新。此外，攻击者可以采用“智能”攻击策略，这种策略的特点是攻击效果不会在单轮训练中显现，而是通过多轮训练的累积效应逐步将后门植入全局模型中。这种场景在实际中是可能存在的，因为中央服务器只能接收到客户端上传的训练后模型更新，而无法知晓每个客户端的具体训练过程。然而，攻击者无法对中央服务器的操作进行干扰，例如，无法更改全局模型的聚合规则或篡改其他良性客户端的本地更新。
\textbf{Attacker's Ability:} The attacker has complete control over the local training process of the malicious robot nodes, meaning they can manipulate local data, trigger patterns, optimization strategies, and local updates at will. Additionally, the attacker may adopt a stealthy attack strategy, where the backdoor is not implanted in a single training round but gradually embedded into the global model through the cumulative effect of multiple rounds of training. This scenario is plausible in practice, as the central server only receives the updates uploaded by robot nodes and has no insight into the specific training process of each node. However, the attacker cannot interfere with the aggregation process of the central server (i.e., altering the global model aggregation rules) or tamper with the local updates of benign nodes.

@@

\subsection{Instantaneous Attack Behavior Perception}
\begin{figure}[!t]
    \centering
    \includegraphics[width=1\linewidth]{figures/secfft-1.pdf}
    \caption{SecFFT's instantaneous attack behavior perception module with frequency-domain deep outlier feature semantics extraction}
    \label{fig2:instant}
\end{figure}
%^\textbf{Motivation:} 目前的防御策略主要集中于利用攻击的“浅层语义特征”来进行恶意节点识别，所谓浅层语义是指模型上传更新的方向\cite{foolsgold}，范数\ref{sun2019can}以及几何中心\cite{pillutla2022robust}等几何或者统计特性。然而，这些基于浅层特征的防御方法在高维空间中存在显著的局限性。首先，这些防御方式只能在特定假设条件下对某些类型的攻击起作用。例如，余弦距离能够检测出具有较大方向偏差的恶意更新，而欧几里得距离则用于检测通过扩大$L_2$范数影响全局模型的恶意更新。同时最新的研究表明\cite{huang2023multi}，基于欧几里得距离等传统的距离度量方法在高维空间中无法有效区分恶意和良性更新，因为这些方法往往受到维度灾难的影响。此外攻击策略也在不停的演变，现在的攻击者已经能够通过精心调整，使恶意更新在方向、范数等浅层特征上与良性更新相似，从而巧妙地避开这些基于浅层语义特征的防御措施，最终在全局模型中植入后门。但尽管攻击者的攻击策略变得越来越复杂，值得注意的是，后门攻击的本质在于让模型建立起预定义触发器和目标之间的联系，因此无论攻击者如何调整自己的攻击方式或策略，其上传的模型更新相较于良性用户始终在某种分布上具有偏移。
\textbf{Motivation:} 
Current defense strategies primarily focus on identifying malicious nodes by leveraging shallow semantic features of the attack, such as the direction \cite{foolsgold}, magnitude \cite{sun2019can}, and geometric or statistical properties like geometric centers \cite{pillutla2022robust} of updates. However, these defense methods based on shallow semantic features face significant limitations in high-dimensional spaces. First, these methods are only effective against certain types of attacks under specific assumptions. For example, cosine distance can detect malicious updates with large directional deviations, while Euclidean distance is used to detect malicious updates that affect the global model by increasing the $L_2$ norm. Furthermore, recent studies \cite{huang2023multi} have shown that traditional distance metrics, such as Euclidean distance, cannot effectively distinguish between malicious and benign updates in high-dimensional spaces due to the curse of dimensionality. Additionally, attack strategies are constantly evolving, and attackers can now carefully adjust their malicious updates to resemble benign updates in terms of direction, magnitude, and other shallow features. This allows them to bypass defenses based on shallow semantic features and eventually implant backdoors into the global model. Despite the increasingly sophisticated strategies employed by attackers, it is important to note that the essence of backdoor attacks lies in establishing a link between the predefined trigger and the target within the model. Thus, regardless of how attackers adjust their attack strategies, their updates will always exhibit some distributional shift compared to benign nodes.

@@

%\textbf{Overview:} 在神经网络中，每个权重代表两个神经元之间连接的强度，权重分布及其相关的能量在训练过程中会经历动态演变，以缩小模型预测结果与真实目标类别之间的差距。当后门攻击迫使神经网络模型学习一种在正常数据中不存在的关联模式时，根据后门触发器的设计和实现方式，这种模式可能是微妙的，也可能是相当明显的，但无论如何，它都代表了与模型从正常数据中学习到的模式的偏离。因此，为解决基于浅层语义特征的防御方法在高维空间中面临的局限性，我们从后门攻击的本质出发\cite{zeng2021rethinking}，提出了一种基于深层语义特征的防御策略，重点分析模型更新在频域空间上的差异。通过将本地模型更新转换到频域，我们能够更准确地捕捉恶意更新在低频成分上的异常表现，这些异常通常难以被传统的浅层特征防御方法发现。通过对频域上的相应特征进行分析，我们可以有效地区分出恶意节点与良性节点，从而增强联邦学习系统在面对复杂后门攻击时的鲁棒性和检测精度。
\textbf{Overview:} 
In neural networks, each weight represents the strength of the connection between two neurons, and the distribution of these weights, along with their associated energy, undergoes dynamic evolution during training to reduce the gap between the model's predicted results and the true classes. nonetheless, When a backdoor attack forces the neural network to learn an association that does not exist in normal data, which may be subtle or quite obvious depending on the design and implementation of the backdoor trigger, it invariably represents a deviation from the association learned by the model from normal data. To address the limitations faced by defense methods based on shallow semantic features in high-dimensional spaces, we propose a new defense strategy rooted in the essence of backdoor attacks \cite{zeng2021rethinking}, which leverages deep semantic features by focusing on the differences in model updates in the frequency domain. According to our experimental results, by transforming local model updates into the frequency domain, we can more accurately capture the abnormal behavior of malicious updates which is typically difficult to detect using traditional shallow feature-based defenses in low-frequency components. By analyzing the relevant features in the frequency domain, we can effectively distinguish between malicious and benign nodes, thereby enhancing the robustness and accuracy of federated learning in the face of sophisticated backdoor attacks.

@@

\subsubsection{Frequency domain transformation}
%受高维空间中维度灾难的影响，传统的基于距离度量的方法已无法有效识别恶意节点。相反，将模型更新转换到频率域后，其差异将主要集中在低频成分上。因此，专注于低频成分的分析，在提升识别准确率的同时，将显著的降低计算开销。假设每个节点上传的更新为 \(\triangledown \theta_i^t \in \mathbb{R}^d\)，其中 \(i\) 表示机器人节点，\(t\) 表示通信轮次，\(d\) 为本地模型上传更新的维度。对于每个更新 \(\triangledown \theta_i^t\)，对其进行归一化，然后通过一维的DCT-II\cite{ahmed1974discrete}变化将其转换为对应的频率分布，并提取$m$个低频分量，例如5000，得到低频向量\(G_i^t\)。
Due to the curse of dimensionality in high-dimensional spaces, traditional methods based on shallow semantic
feature can no longer effectively identify malicious nodes. In contrast, once model updates are transformed into the frequency domain, their differences are primarily concentrated in the low-frequency components. Focusing on the analysis of these low-frequency components not only improves identification accuracy but also significantly reduces computational overhead. For each update \(\triangledown \theta_i^t\), we apply one-dimensional DCT-II \cite{ahmed1974discrete} to obtain its corresponding frequency distribution. Then we extract $m$ low-frequency components\cite{fereidooni2023freqfed}, such as 5000, yielding the low-frequency vector $G_i^t$.

@@

\subsubsection{Clean ingredient extraction}
% 在提取到节点更新对应的低频向量\(G_i^t\)后，我们使用流行的聚类算法例如HDBSCAN\cite{mcinnes2017hdbscan}对这些更新进行聚类，得到多个簇$C_0^t, C_1^t, \ldots, C_{\kappa}^t$，其中$\kappa$表示簇的数量。在大多数联邦学习场景中，良性节点的数量通常多于恶意节点的数量，且良性节点的更新通常更接近全局模型的方向并表现出更高的聚集性。基于这一观察，我们选择节点数量最多的簇$C_{max} = \{R_{i_0}, R_{i_1}, \ldots, R_{i_n}\}$作为主要分析对象。为了进一步提取该簇的主要特征或模式，我们对其中所有节点组成的矩阵进行奇异值分解\cite{wang2023scfl}。由于，SVD分解的左奇异向量$\tilde{G}^t$对应于数据的主要方向，因此，$\tilde{G}^t$可以作为良性节点在低频域内的主要特征。
After extracting the corresponding low-frequency vectors \(G_i^t\) from updates, we employ popular clustering algorithms such as HDBSCAN \cite{mcinnes2017hdbscan} to cluster these vectors, resulting in multiple clusters \(C_0^t, C_1^t, \ldots, C_{\kappa}^t\), where \(\kappa\) denotes the number of clusters. In most federated learning scenarios, the number of benign nodes typically exceeds that of malicious nodes, which is consistent with the assumptions outlined in our threat model. Moreover, the updates from benign nodes are generally closer to the direction of the global model and exhibit a higher degree of aggregation. This behavior is also evident in the frequency domain, where it becomes even more pronounced. Based on this observation, we select the largest cluster \(C^t_{max} = \{G^t_{i_0}, G^t_{i_1}, \ldots, G^t_{i_n}\}\) as the primary subject for analysis. To further extract the main characteristics of this cluster, we perform singular value decomposition (SVD)\cite{wang2023scfl} on the matrix $H^t = (G^t_{i_0}, G^t_{i_1}, \ldots, G^t_{i_n}) \in \mathbb{R}^{m \times n}$ formed by all nodes in this cluster. Since the left singular vectors of the SVD, denoted as \(\tilde{G}^t\), represent the principal directions of these vectors, \(\tilde{G}^t\) can be regarded as the principal ingredient of benign nodes in the low-frequency domain.

@@

\begin{equation}
Chi^t_i = \sqrt{ \sum_{k=0}^{m-1}\frac{(G_i^t[k] - {\tilde G}^t[k])^2}{{|\tilde G}^t[k]| + \epsilon} }
\label{equation:KL}
\end{equation}
% 其中 \(m\) 是低频向量的维度，\(\epsilon\) 是一个小的常数，用于避免分母为零。相比于传统的欧氏距离，卡方距离根据两个向量对应分量的大小进行加权，使得特征值较大的部分对距离的影响更大，从而能够更准确地反映低频向量中重要特征之间的差异，同时忽略不重要的噪声成分。
where \(G_i^t[k]\) denotes the k-th element of vector \(G_i\) and \(\epsilon\) is a small constant to avoid division by zero.

Chi-square distance is usually used to measure whether a certain observed distribution conforms to a typical theoretical distribution, which is very consistent with our method because we use the calculated clean ingredient as a criterion for measuring benign and malicious nodes. Compared with the traditional Euclidean distance, the chi-square distance assigns weights according to the clean ingredient, giving greater importance to smaller components. On the other hand, cosine similarity measures the angle between two vectors and is usually used to evaluate their directional consistency, ignoring magnitude differences. However, in our method, we are concerned with the degree of deviation of each user's low-frequency vector $G_i^t$ from the clean ingredient $\tilde{G}^t$, including magnitude differences, so the chi-square distance is more suitable for capturing hidden malicious attacks than cosine similarity.

% 为什么要用卡方距离，不用欧式距离？实验论证？

@@

In our proposed framework, the concept of the "intention point" serves as a pivotal element in recognizing the underlying motives of malicious nodes during the federated fine-tuning process. The intention point is defined as the central position within a high-dimensional space, representing the aggregated gradient updates of a node across multiple training rounds. This point encapsulates the node's intended direction for influencing the global model. To construct the intention point, we utilize a Minimum Enclosing Hypersphere approach, which encompasses a proportion of the node’s gradient updates. The radius of this hypersphere reflects the reliability of the intention point, effectively quantifying the node's consistency in updating the model. By analyzing these intention points, our method can discern the behaviors of malicious nodes from benign ones, enhancing the overall security and robustness of the federated learning system.
